Swinging, Fast and Slow

### 2024-08-24 reports/saberseminar/slides.tex
continued march toward saberseminar

\usepackage{hyperref, tikz}
\usetikzlibrary{arrows.meta}
\tikzset{>={To[length=3mm,width=3mm]}}

@@

@@

\begin{frame}{Caution \#2: How do we get our measurement?}
  \includegraphics[width = 0.49\textwidth]{../../figures/beamer/swing_late.pdf}
  \includegraphics[width = 0.49\textwidth]{../../figures/beamer/swing_early.pdf}
  \begin{itemize}
    \item For \alert{identical} swings, timing determines point of measurement
  \end{itemize}
\end{frame}

\begin{frame}{Outline}
  \begin{tikzpicture}[
    ->,
    align = center,
    minimum height = 1.5cm,
    minimum width = 2cm,
    alert/.style = {thick, draw = solarizedAccent}
  ]
    \node[draw] (intention_model) at (0, 0) {Intention\\Model};
    \node[draw] (approach) at (3, 2.5) {Approach};
    \node[draw] (adaptation) at (3, 0) {Adaptation};
    \node[draw] (timing) at (3, -2.5) {Timing};
    \node[draw] (pitch_outcome_model) at (6, 0) {Pitch\\Outcome\\Model};
    \node[draw] (approach_model) at (6, 2.5) {Approach\\Model};
    \node[draw] (count_value) at (9, 0) {Count\\Value};
    \node[draw] (approach_value) at (9, 2.5) {Approach\\Value};
    \draw (intention_model) -- (approach);
    \draw (intention_model) -- (adaptation);
    \draw (intention_model) -- (timing);
    \draw (pitch_outcome_model) -- (approach_model);
    \draw (approach) -- (approach_model);
    \draw (count_value) -- (approach_value);
    \draw (approach_model) -- (approach_value);
  \end{tikzpicture}
\end{frame}

\begin{frame}{Outline}
  \begin{tikzpicture}[
    ->,
    align = center,
    minimum height = 1.5cm,
    minimum width = 2cm,
    alert/.style = {thick, draw = solarizedAccent}
  ]
    \node[alert] (intention_model) at (0, 0) {\alert{Intention}\\\alert{Model}};
    \node[draw] (approach) at (3, 2.5) {Approach};
    \node[draw] (adaptation) at (3, 0) {Adaptation};
    \node[draw] (timing) at (3, -2.5) {Timing};
    \node[draw] (pitch_outcome_model) at (6, 0) {Pitch\\Outcome\\Model};
    \node[draw] (approach_model) at (6, 2.5) {Approach\\Model};
    \node[draw] (count_value) at (9, 0) {Count\\Value};
    \node[draw] (approach_value) at (9, 2.5) {Approach\\Value};
    \draw (intention_model) -- (approach);
    \draw (intention_model) -- (adaptation);
    \draw (intention_model) -- (timing);
    \draw (pitch_outcome_model) -- (approach_model);
    \draw (approach) -- (approach_model);
    \draw (count_value) -- (approach_value);
    \draw (approach_model) -- (approach_value);
  \end{tikzpicture}
\end{frame}

\begin{frame}{Intention Model}
  Goal: Estimate swing metrics at intended point of contact
  \begin{enumerate}
    \item Filter on swings above 50 mph (avoid bunts, check-swings)
    \item Filter on swings that are squared-up (avoid bad timing)
    \item Fit mixed-effects linear models for swing length and bat speed
    \begin{itemize}
      \item Fixed and random effects for intercept, count, pitch location
    \end{itemize}
  \end{enumerate}
  \begin{align*}
    (\mbox{swing length})_i = \alpha &+ \gamma^A_{b_i}\\
        & \left.
          \begin{array}{l}
            +~\beta^B \cdot (\mbox{balls})_i\\
            +~(\beta^S + \gamma^S_{b_i}) \cdot (\mbox{strikes})_i
          \end{array}\color{solarizedAccent}
        \right\}\mbox{\alert{approach}}\\
        & \left.
          \begin{array}{l}
            +~(\beta^X + \gamma^X_{b_i}) \cdot (\mbox{pitch loc x})_i\\
            +~(\beta^Z + \gamma^Z_{b_i}) \cdot (\mbox{pitch loc z})_i
          \end{array}\color{solarizedAccent}
        \right\}\mbox{\alert{adaptation}}\\
        & \left.
          \begin{array}{l}
            +~\epsilon_i
          \end{array}
        \right.\color{solarizedAccent}\leftarrow\mbox{\alert{timing}}
  \end{align*}
\end{frame}

@@

\begin{frame}{Outline}
  \begin{tikzpicture}[
    ->,
    align = center,
    minimum height = 1.5cm,
    minimum width = 2cm,
    alert/.style = {thick, draw = solarizedAccent}
  ]
    \node[draw] (intention_model) at (0, 0) {Intention\\Model};
    \node[draw] (approach) at (3, 2.5) {Approach};
    \node[draw] (adaptation) at (3, 0) {Adaptation};
    \node[draw] (timing) at (3, -2.5) {Timing};
    \node[draw] (pitch_outcome_model) at (6, 0) {Pitch\\Outcome\\Model};
    \node[alert] (approach_model) at (6, 2.5) {\alert{Approach}\\\alert{Model}};
    \node[draw] (count_value) at (9, 0) {Count\\Value};
    \node[draw] (approach_value) at (9, 2.5) {Approach\\Value};
    \draw (intention_model) -- (approach);
    \draw (intention_model) -- (adaptation);
    \draw (intention_model) -- (timing);
    \draw (pitch_outcome_model) -- (approach_model);
    \draw (approach) -- (approach_model);
    \draw (count_value) -- (approach_value);
    \draw (approach_model) -- (approach_value);
  \end{tikzpicture}
\end{frame}

\begin{frame}{Approach Model}
  \begin{enumerate}
    \item Start with pitch outcome model which estimates $\mathbb{P}$(swing), $\mathbb{P}$(contact $\mid$ swing), $\mathbb{P}$(fair $\mid$ contact), $\mathbb{E}$(xwOBA $\mid$ fair), etc.
    \item Refit contact, fair and xwOBA models on batter approach, with previous predictions as offset
  \end{enumerate}
  Example: $p_i$ is contact prob, $\hat p_i$ is pitch outcome model estimate
  \begin{align*}
    \log\left(\frac{p_i}{1 - p_i}\right) = \log\left(\frac{\hat p_i}{1 - \hat p_i}\right) + \beta \cdot \hat\gamma^S_{b_i} \cdot (\mbox{strikes})_i
  \end{align*}
  \begin{itemize}
    \item This is instrumental variable regression from causal inference, using count as the instrument
  \end{itemize}
\end{frame}

@@

\begin{frame}{Outline}
  \begin{tikzpicture}[
    ->,
    align = center,
    minimum height = 1.5cm,
    minimum width = 2cm,
    alert/.style = {thick, draw = solarizedAccent}
  ]
    \node[draw] (intention_model) at (0, 0) {Intention\\Model};
    \node[draw] (approach) at (3, 2.5) {Approach};
    \node[draw] (adaptation) at (3, 0) {Adaptation};
    \node[draw] (timing) at (3, -2.5) {Timing};
    \node[draw] (pitch_outcome_model) at (6, 0) {Pitch\\Outcome\\Model};
    \node[draw] (approach_model) at (6, 2.5) {Approach\\Model};
    \node[draw] (count_value) at (9, 0) {Count\\Value};
    \node[alert] (approach_value) at (9, 2.5) {\alert{Approach}\\\alert{Value}};
    \draw (intention_model) -- (approach);
    \draw (intention_model) -- (adaptation);
    \draw (intention_model) -- (timing);
    \draw (pitch_outcome_model) -- (approach_model);
    \draw (approach) -- (approach_model);
    \draw (count_value) -- (approach_value);
    \draw (approach_model) -- (approach_value);
  \end{tikzpicture}
\end{frame}

@@

>>>>

\usepackage{hyperref, tikz}
\usetikzlibrary{arrows.meta, shapes.geometric}
\tikzset{>={To[length=3mm,width=3mm]}}

@@

\begin{frame}{Question}
  How much do batters change their swings according to the count?\\
  ~\\
  And what are the consequences of these approaches?
\end{frame}

@@

\begin{frame}{Caution \#2: How do we get our measurement?}
  \includegraphics[width = 0.49\textwidth]{../../figures/beamer/swing_late.pdf}
  \includegraphics[width = 0.49\textwidth]{../../figures/beamer/swing_early.pdf}
  \begin{itemize}
    \item For \alert{identical} swings, timing determines point of measurement
    \item Swing length and bat speed are \alert{outcomes}, not purely process
  \end{itemize}
\end{frame}

\begin{frame}{Outline}
  \begin{tikzpicture}[
    ->,
    align = center,
    minimum height = 1.5cm,
    minimum width = 2cm,
    every node/.style = {draw},
    model/.style = {rectangle},
    value/.style = {ellipse},
    alert/.style = {thick, draw = solarizedAccent}
  ]
    \node[model] (intention_model) at (0, 0) {Intention\\Model};
    \node[value] (approach) at (3, 2.5) {Approach};
    \node[value] (adaptation) at (3, 0) {Adaptation};
    \node[value] (timing) at (3, -2.5) {Timing};
    \node[model] (pitch_outcome_model) at (6, 0) {Pitch\\Outcome\\Model};
    \node[model] (approach_model) at (6, 2.5) {Approach\\Model};
    \node[model] (markov_model) at (9, 0) {Markov\\Model\\for PA};
    \node[value] (approach_value) at (9, 2.5) {Approach\\Value};
    \draw (intention_model) -- (approach);
    \draw (intention_model) -- (adaptation);
    \draw (intention_model) -- (timing);
    \draw (pitch_outcome_model) -- (approach_model);
    \draw (approach) -- (approach_model);
    \draw (markov_model) -- (approach_value);
    \draw (approach_model) -- (approach_value);
  \end{tikzpicture}
\end{frame}

\begin{frame}{Outline}
  \begin{tikzpicture}[
    ->,
    align = center,
    minimum height = 1.5cm,
    minimum width = 2cm,
    every node/.style = {draw},
    model/.style = {rectangle},
    value/.style = {ellipse},
    alert/.style = {thick, draw = solarizedAccent}
  ]
    \node[model, alert] (intention_model) at (0, 0) {\alert{Intention}\\\alert{Model}};
    \node[value] (approach) at (3, 2.5) {Approach};
    \node[value] (adaptation) at (3, 0) {Adaptation};
    \node[value] (timing) at (3, -2.5) {Timing};
    \node[model] (pitch_outcome_model) at (6, 0) {Pitch\\Outcome\\Model};
    \node[model] (approach_model) at (6, 2.5) {Approach\\Model};
    \node[model] (markov_model) at (9, 0) {Markov\\Model\\for PA};
    \node[value] (approach_value) at (9, 2.5) {Approach\\Value};
    \draw (intention_model) -- (approach);
    \draw (intention_model) -- (adaptation);
    \draw (intention_model) -- (timing);
    \draw (pitch_outcome_model) -- (approach_model);
    \draw (approach) -- (approach_model);
    \draw (markov_model) -- (approach_value);
    \draw (approach_model) -- (approach_value);
  \end{tikzpicture}
\end{frame}

\begin{frame}{Intention Model}
  Goal: Estimate swing metrics at intended point of contact
  \begin{enumerate}
    \item Filter on swings above 50 mph (avoid bunts, check-swings)
    \item Filter on swings that are squared-up (avoid bad timing)
    \item Fit mixed-effects linear models for swing length and bat speed
    \begin{itemize}
      \item Fixed and random effects for intercept, count, pitch location
    \end{itemize}
  \end{enumerate}
  \begin{align*}
    (\mbox{swing length})_i = \alpha &+ \gamma^A_{b_i}\\
        & \left.
          \begin{array}{l}
            +~\beta^B \cdot (\mbox{balls})_i\\
            +~(\beta^S + \gamma^S_{b_i}) \cdot (\mbox{strikes})_i
          \end{array}\color{solarizedAccent}
        \right\}\mbox{\alert{approach}}\\
        & \left.
          \begin{array}{l}
            +~(\beta^X + \gamma^X_{b_i}) \cdot (\mbox{pitch loc x})_i\\
            +~(\beta^Z + \gamma^Z_{b_i}) \cdot (\mbox{pitch loc z})_i
          \end{array}\color{solarizedAccent}
        \right\}\mbox{\alert{adaptation}}\\
        & \left.
          \begin{array}{l}
            +~\epsilon_i
          \end{array}
        \right.
  \end{align*}
\end{frame}

@@

\begin{frame}{Outline}
  \begin{tikzpicture}[
    ->,
    align = center,
    minimum height = 1.5cm,
    minimum width = 2cm,
    every node/.style = {draw},
    model/.style = {rectangle},
    value/.style = {ellipse},
    alert/.style = {thick, draw = solarizedAccent}
  ]
    \node[model] (intention_model) at (0, 0) {Intention\\Model};
    \node[value] (approach) at (3, 2.5) {Approach};
    \node[value] (adaptation) at (3, 0) {Adaptation};
    \node[value] (timing) at (3, -2.5) {Timing};
    \node[model] (pitch_outcome_model) at (6, 0) {Pitch\\Outcome\\Model};
    \node[model, alert] (approach_model) at (6, 2.5) {\alert{Approach}\\\alert{Model}};
    \node[model] (markov_model) at (9, 0) {Markov\\Model\\for PA};
    \node[value] (approach_value) at (9, 2.5) {Approach\\Value};
    \draw (intention_model) -- (approach);
    \draw (intention_model) -- (adaptation);
    \draw (intention_model) -- (timing);
    \draw (pitch_outcome_model) -- (approach_model);
    \draw (approach) -- (approach_model);
    \draw (markov_model) -- (approach_value);
    \draw (approach_model) -- (approach_value);
  \end{tikzpicture}
\end{frame}

\begin{frame}{Approach Model}
  \begin{enumerate}
    \item Start with pitch outcome model which estimates $\mathbb{P}$(swing), $\mathbb{P}$(contact $\mid$ swing), $\mathbb{P}$(fair $\mid$ contact), $\mathbb{E}$(xwOBA $\mid$ fair), etc.
    \begin{itemize}
      \item Gradient boosting w/ 2022-2024 MLB data
    \end{itemize}
    \item Refit contact, fair and xwOBA models on batter approach, with previous predictions as offset
  \end{enumerate}
  Example: $p_i$ is contact prob, $\hat p_i$ is pitch outcome model estimate
  \begin{align*}
    \log\left(\frac{p_i}{1 - p_i}\right) = \log\left(\frac{\hat p_i}{1 - \hat p_i}\right) + \beta \cdot \hat\gamma^S_{b_i} \cdot (\mbox{strikes})_i
  \end{align*}
  \begin{itemize}
    \item This is instrumental variable regression from causal inference, using count as the instrument
  \end{itemize}
\end{frame}

@@

\begin{frame}{Outline}
  \begin{tikzpicture}[
    ->,
    align = center,
    minimum height = 1.5cm,
    minimum width = 2cm,
    every node/.style = {draw},
    model/.style = {rectangle},
    value/.style = {ellipse},
    alert/.style = {thick, draw = solarizedAccent}
  ]
    \node[model] (intention_model) at (0, 0) {Intention\\Model};
    \node[value] (approach) at (3, 2.5) {Approach};
    \node[value] (adaptation) at (3, 0) {Adaptation};
    \node[value] (timing) at (3, -2.5) {Timing};
    \node[model] (pitch_outcome_model) at (6, 0) {Pitch\\Outcome\\Model};
    \node[model] (approach_model) at (6, 2.5) {Approach\\Model};
    \node[model, alert] (markov_model) at (9, 0) {\alert{Markov}\\\alert{Model}\\\alert{for PA}};
    \node[value] (approach_value) at (9, 2.5) {Approach\\Value};
    \draw (intention_model) -- (approach);
    \draw (intention_model) -- (adaptation);
    \draw (intention_model) -- (timing);
    \draw (pitch_outcome_model) -- (approach_model);
    \draw (approach) -- (approach_model);
    \draw (markov_model) -- (approach_value);
    \draw (approach_model) -- (approach_value);
  \end{tikzpicture}
\end{frame}

@@

\begin{frame}{Is it good to modulate your swing by count?}
  \centering
  \includegraphics[height = 0.9\textheight]{../../figures/beamer/approach_run_value_labelled.pdf}
\end{frame}
